---
title: "Summary statistics for treatment efficacy estimation
  in clinical trials: when simplicity rivals complexity"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  knitr.table.format = "latex",
  scipen = 1,
  digits = 3
)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(tidyr)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  results = "asis",
  dev = "pdf",
  cache = TRUE,
  cache.path = "cache/"
)
```

# Introduction

Mixed-model repeated measures (MMRM) has become the default
analytic strategy for longitudinal clinical trials with
continuous endpoints. Regulatory agencies and methodological
guidance documents increasingly endorse MMRM as the primary
analysis [@ichE9R12019], and the approach has clear theoretical
advantages: it uses all available data under the missing-at-random
(MAR) assumption, accommodates unbalanced designs, and models
complex within-subject correlation structures
[@lairdRandomEffectsModels1982].

Yet MMRM carries substantial practical burdens. The method
requires specification of a covariance structure, and
misspecification can inflate Type I error or reduce power
[@vossoughiSummaryMeasureAnalysis2012]. Degrees-of-freedom
approximations---Satterthwaite versus Kenward-Roger---remain
unsettled, with different software implementations yielding
discrepant p-values for the same data
[@kenwardSmallSampleInference1997; @mmrmPackage2024].
Convergence failures are common in R packages when fitting
unstructured covariance matrices, particularly with small
samples or many time points. And perhaps most critically, the
MMRM treatment effect estimate at a single visit is not always
intuitive to clinical investigators, who may prefer a single,
interpretable summary of each patient's response trajectory.

An older and simpler tradition---the summary measures
approach---offers a compelling alternative. First articulated
systematically by @matthewsAnalysisSerialMeasurements1990,
this two-stage method reduces each subject's repeated
measurements to a single scalar summary (e.g., a slope, area
under the curve, post-treatment mean, or change score) and
then applies standard between-group tests (t-tests, ANCOVA)
to these summaries. The approach requires no covariance
structure specification, yields exact p-values from
well-understood distributions, and produces treatment effect
estimates that are directly interpretable.

The question addressed in the present paper is whether there
exist clinical trial settings in which summary statistic-based
inference is as effective as or superior to MMRM, and what
those settings look like.

## The summary measures tradition

The intellectual history of summary measures in clinical trials
spans three decades. @matthewsAnalysisSerialMeasurements1990
introduced the two-stage framework in the BMJ, arguing that the
then-common practice of testing group differences at each time
point was both statistically invalid (due to multiplicity) and
clinically uninformative. They proposed that investigators first
identify a summary that captures the clinically relevant feature
of each patient's trajectory, then analyse these summaries with
simple methods. The approach is statistically valid because the
summaries, being one per subject, are independent across
subjects.

@frisonRepeatedMeasuresClinical1992 extended this framework
substantially by demonstrating that when the data can be
summarized by pre-treatment and post-treatment means, analysis
of covariance (ANCOVA) of the post-treatment means adjusted for
the pre-treatment means is the method of choice. They quantified
the superiority of ANCOVA over analysis of post-treatment means
alone and over analysis of change scores, showing that ANCOVA
dominates both in terms of reduced variance and avoidance of
bias. Importantly, they showed that having more than one
pre-treatment measurement provides meaningful gains in
precision, because the pre-treatment mean is a better estimate
of the true baseline than a single observation.

@frisonLinearlyDivergentTreatment1997 addressed the case where
treatment effects diverge linearly over time. For this setting,
the natural summary statistic is each subject's regression slope,
and the authors showed that adjusting slopes for baseline values
(or better, for estimated intercepts) yields substantial
efficiency gains. They derived the optimal linear summary
statistic for any repeated measures design with known covariance
structure and known shape of the mean treatment difference,
demonstrating that the summary measures approach can be made
nearly fully efficient.

@sennRepeatedMeasuresClinical2000 provided a practical
synthesis, arguing that much repeated measures analysis using
mixed models could be replaced by a summary measures approach
with negligible loss of efficiency. Senn and colleagues described
strategies for baseline adjustment with multiple baselines and
proposed a compromise trend/mean measure (regression through the
origin) useful when both the level and rate of change matter.
They also made the provocative observation that some repeated
measures approaches implicitly violate intention-to-treat
principles, a problem that summary measures avoid.

## Extensions and refinements

@dawsonStratificationSummaryStatistic1994 recognized that
summary statistic distributions depend on the pattern of
missingness and proposed stratifying summary statistic tests
by missing data pattern. Simulations showed that stratification
increases power and improves robustness to violations of
missing data assumptions. @dawsonSampleSizeCalculations1998
later provided a unified sample size framework for summary
statistics, covering slopes, post-baseline means, change
scores, and final observations, with modifications for missing
data from staggered entry and random dropouts.

@weinbergEfficiencyComparisonsRank2001 examined the efficiency
of rank and permutation tests based on summary statistics under
more complex distributional assumptions, including censored
laboratory markers. Their work demonstrated that the choice of
summary statistic can substantially affect asymptotic relative
efficiency, particularly when the repeated measures exhibit
nonlinear growth or left censoring.

@sennChangeBaselineAnalysis2006 revisited the choice between
ANCOVA and change-score analysis, clarifying persistent
confusion in the literature. Senn argued that ANCOVA is
generally preferable for randomized trials because it adjusts
for chance imbalances in baseline values without introducing
the bias that afflicts change-score analysis in non-randomized
settings.

## MMRM: strengths and practical limitations

The theoretical foundation for MMRM rests on the random-effects
model of @lairdRandomEffectsModels1982, which provides maximum
likelihood and restricted maximum likelihood (REML) estimation
for models with structured covariance. The key practical
advantage is that MMRM uses all available data, providing valid
inference under MAR without requiring imputation.

@mallinckrodtEfficacyDuloxetine2004 provided an influential
empirical comparison of MMRM and LOCF-ANCOVA across eight
duloxetine trials. While MMRM and LOCF agreed on significance
in 82\% of 202 comparisons, MMRM yielded significant results
in 12.4\% of cases where LOCF did not. This work cemented
MMRM's position as the preferred analysis but also revealed
that the two approaches diverge substantively in a non-trivial
minority of cases.

However, MMRM faces several practical limitations:

\begin{enumerate}
\item \textbf{Degrees-of-freedom approximations.}
  @kenwardSmallSampleInference1997 proposed a scaled Wald
  statistic with an $F$ approximation that reduces
  small-sample bias, but the Kenward-Roger adjustment depends
  on the parameterization of the covariance model, producing
  different results across software implementations. The
  alternative Satterthwaite approximation is computationally
  simpler but can be liberal in small samples. Neither method
  has a clear theoretical advantage in all settings.

\item \textbf{Software fragmentation.} SAS \texttt{PROC MIXED}
  has long been the reference implementation, but R packages
  (\texttt{nlme}, \texttt{lme4}/\texttt{lmerTest},
  \texttt{glmmTMB}) have exhibited convergence problems,
  particularly with unstructured covariance matrices. The
  \texttt{mmrm} package [@mmrmPackage2024] was developed
  specifically to address these gaps, but discrepancies
  between SAS and R results persist due to different
  parameterizations.

\item \textbf{Covariance structure specification.}
  @vossoughiSummaryMeasureAnalysis2012 showed that
  misspecification of the error covariance structure in LMM
  can produce non-sensible results, whereas the summary
  measure approach is robust to the true covariance structure.
  In practice, analysts often default to unstructured
  covariance, but this requires estimation of $p(p+1)/2$
  parameters for $p$ time points, which may exceed the
  information in the data.

\item \textbf{Interpretability.} The MMRM treatment effect at
  a specific visit (e.g., week 24) is a conditional quantity
  that depends on the covariance model. The analogous summary
  statistic (e.g., slope difference) has a direct clinical
  interpretation: how much faster does the treatment group
  improve per unit time?

\item \textbf{Time-by-covariate interactions.}
  @liuMixedModelsCovariateInteractions2021 demonstrated that
  MMRM must include time-by-covariate interactions to achieve
  power gains and robustness against dropout bias relative to
  complete-case ANCOVA. Without these interactions, MMRM may
  not outperform a simple ANCOVA of the final visit.
\end{enumerate}

## When are summary statistics adequate?

The performance comparison literature provides guidance on when
summary statistics match or exceed MMRM.
@vossoughiSummaryMeasureAnalysis2012 conducted extensive Monte
Carlo simulations comparing the summary measure approach (SMA),
linear mixed models (LMM), and the unstructured multivariate
approach (UMA) for linear trend data. They found that SMA
completely dominated UMA and performed ``convincingly close''
to the best-fitting LMM in testing all effects. Critically,
when the LMM covariance structure was misspecified, the LMM
produced unreliable results, while SMA remained robust. They
concluded that SMA is ``a simple, safe and powerful approach in
which the loss of efficiency compared to the best-fitting LMM
was generally negligible'' and recommended it as ``the first
choice to reliably analyse the linear trend data with a
moderate to large number of measurements and/or small to
moderate sample sizes.''

In the Alzheimer's disease trial literature,
@ardMixedModelRepeated2011 compared MMRM with slope models
for cognitive outcomes and found that slope models provide a
more intuitive and clinically meaningful way of demonstrating
disease-modifying effects. @ardSimulationStudyComparing2015
followed with a simulation study showing that both methods
maintain proper Type I error, but the slope model has a
moderate power advantage when the true progression is linear
or near-linear. When disease progression is nonlinear, MMRM
with its semiparametric treatment of time retains an advantage.

Based on the theoretical and empirical literature, summary
statistics appear to be as effective as MMRM in the following
settings:

\begin{itemize}
\item Treatment effects are expected to diverge linearly or
  to reach a stable plateau (constant treatment effect after
  an initial period).
\item The number of repeated measurements is moderate to
  large (enabling stable estimation of per-subject summaries).
\item Missing data rates are low to moderate ($<$20\%) and
  are not strongly informative.
\item Sample sizes are small to moderate, where MMRM's
  degrees-of-freedom approximations are least reliable.
\item Multiple pre-treatment measurements are available,
  enabling precise baseline adjustment.
\item The scientific question concerns the overall rate of
  change or cumulative treatment benefit, rather than the
  effect at a single specific visit.
\end{itemize}

## Present study

The objectives of the present study are threefold: (1) to
synthesize the theoretical and empirical literature comparing
summary statistic and MMRM approaches; (2) to identify
specific clinical trial design features that favour summary
statistics; and (3) to conduct a simulation study comparing
the operating characteristics of summary statistic-based
ANCOVA and MMRM across a range of design parameters relevant
to clinical practice.

# Methods

## Data generating model

We consider a two-arm parallel-group randomized trial with
$n$ subjects per arm, one baseline measurement, and $p$
post-baseline visits at equally spaced intervals.
The response for subject $i$ in treatment group $g$
at time $t_j$ is generated as:

$$Y_{igj} = \beta_0 + b_{0i} + (\beta_1 + b_{1i}) t_j
  + \delta_g \cdot t_j + \epsilon_{igj}$$

where $\beta_0$ is the overall intercept, $b_{0i} \sim
N(0, \sigma^2_{b0})$ is the random intercept, $\beta_1$ is
the overall slope, $b_{1i} \sim N(0, \sigma^2_{b1})$ is the
random slope, $\delta_g$ is the treatment effect on the slope
($\delta_0 = 0$ for placebo), and $\epsilon_{igj} \sim
N(0, \sigma^2_\epsilon)$ is residual error. Random effects
are assumed independent of residual errors, and
$\text{Cor}(b_{0i}, b_{1i}) = \rho_{01}$.

## Parameter values

```{r sim-params}
params <- tibble::tibble(
  Parameter = c(
    '$n$ (per arm)', '$p$ (post-baseline visits)',
    '$\\beta_0$ (intercept)', '$\\beta_1$ (placebo slope)',
    '$\\delta$ (treatment slope effect)',
    '$\\sigma_{b0}$ (random intercept SD)',
    '$\\sigma_{b1}$ (random slope SD)',
    '$\\rho_{01}$ (intercept-slope correlation)',
    '$\\sigma_\\epsilon$ (residual SD)',
    'Missing data rate'
  ),
  Values = c(
    '30, 50, 100', '4, 8',
    '50', '-0.5',
    '0.25',
    '5', '0.3',
    '-0.3',
    '3',
    '0\\%, 10\\%, 20\\%'
  )
)

kable(
  params,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = "Simulation parameter values."
) |>
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  )
```

## Design specifications

We compare three analytic approaches:

\begin{enumerate}
\item \textbf{Slope summary + ANCOVA (SMA-slope).}
  For each subject, fit a simple linear regression of
  post-baseline responses on time. The per-subject slope is
  the summary statistic. Compare treatment groups using
  ANCOVA with the baseline measurement as covariate.

\item \textbf{Change score + ANCOVA (SMA-change).}
  For each subject, compute the mean of all post-baseline
  measurements minus the baseline measurement. Compare
  treatment groups using a two-sample t-test (equivalent
  to ANCOVA without baseline adjustment) and ANCOVA with
  baseline adjustment.

\item \textbf{MMRM.} Fit a mixed-effects model with
  fixed effects for treatment, visit (categorical), and
  treatment-by-visit interaction, with baseline as covariate.
  Use unstructured covariance and Satterthwaite
  degrees of freedom. The treatment effect is tested at the
  final visit.
\end{enumerate}

## Missing data mechanism

Missing data are generated under two mechanisms:

- \textbf{MCAR}: each post-baseline observation is
  independently missing with probability equal to the
  specified rate.
- \textbf{MAR}: the probability of dropout at visit $j$
  depends on the observed response at visit $j-1$, with
  subjects showing worse outcomes more likely to drop out.

## Simulation procedure

```{r sim-source, cache=FALSE}
sim_script <- file.path(
  here::here(), "analysis", "scripts", "sim_study.R"
)
if (file.exists(sim_script)) {
  source(sim_script)
}
```

```{r run-sim, cache=TRUE, eval=FALSE}
set.seed(20260310)
n_reps <- 1000

design_grid <- expand.grid(
  n_per_arm = c(30, 50, 100),
  p_visits = c(4, 8),
  miss_rate = c(0, 0.10, 0.20),
  miss_mech = c("MCAR", "MAR")
)

results <- purrr::pmap_dfr(
  design_grid,
  function(n_per_arm, p_visits, miss_rate, miss_mech) {
    run_simulation(
      n_per_arm = n_per_arm,
      p_visits = p_visits,
      miss_rate = miss_rate,
      miss_mech = miss_mech,
      n_reps = n_reps
    )
  }
)
```

For each of the 12 design conditions
and each of 1,000 replications, we record: the point estimate
of the treatment effect, its standard error, the p-value, and
whether the 95\% confidence interval covers the true value.

## Performance metrics

For each design condition and analytic method, we compute:

- **Bias**: mean estimate minus true treatment effect
- **Empirical SE**: standard deviation of estimates across
  replications
- **Mean model-based SE**: mean of estimated standard errors
- **Power**: proportion of replications with $p < 0.05$
- **Coverage**: proportion of 95\% CIs containing the true
  value
- **Convergence rate**: proportion of replications where the
  model fit converged (relevant for MMRM only)

# Results

## Summary statistics

*[Results to be populated after simulation is executed.]*

## Figures

```{r power-plot, eval=FALSE, fig.cap="Power comparison across methods and design conditions.", fig.width=6, fig.height=4}
results |>
  filter(miss_mech == "MCAR") |>
  ggplot(aes(
    x = factor(n_per_arm),
    y = power,
    shape = method,
    linetype = method,
    group = method
  )) +
  geom_point(size = 2) +
  geom_line() +
  facet_grid(
    miss_rate ~ p_visits,
    labeller = labeller(
      miss_rate = function(x) paste0(
        as.numeric(x) * 100, "% missing"
      ),
      p_visits = function(x) paste0(x, " visits")
    )
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),
    name = "Method"
  ) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted"),
    name = "Method"
  ) +
  labs(
    x = "Sample size per arm",
    y = "Power"
  ) +
  theme_minimal(base_size = 10)
```

```{r coverage-plot, eval=FALSE, fig.cap="95\\% CI coverage across methods and design conditions.", fig.width=6, fig.height=4}
results |>
  filter(miss_mech == "MCAR") |>
  ggplot(aes(
    x = factor(n_per_arm),
    y = coverage,
    shape = method,
    linetype = method,
    group = method
  )) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(
    yintercept = 0.95,
    linetype = "dashed",
    colour = "grey50"
  ) +
  facet_grid(
    miss_rate ~ p_visits,
    labeller = labeller(
      miss_rate = function(x) paste0(
        as.numeric(x) * 100, "% missing"
      ),
      p_visits = function(x) paste0(x, " visits")
    )
  ) +
  scale_shape_manual(
    values = c(16, 17, 15),
    name = "Method"
  ) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotted"),
    name = "Method"
  ) +
  labs(
    x = "Sample size per arm",
    y = "Coverage"
  ) +
  theme_minimal(base_size = 10)
```

# Discussion

The central thesis of this paper is that the summary measures
approach deserves reconsideration as a primary analysis for
certain classes of clinical trials. The theoretical literature,
beginning with @matthewsAnalysisSerialMeasurements1990 and
formalized by @frisonRepeatedMeasuresClinical1992 and
@frisonLinearlyDivergentTreatment1997, establishes that
summary statistics with ANCOVA baseline adjustment can achieve
near-optimal efficiency for detecting linear treatment effects.
The simulation work of @vossoughiSummaryMeasureAnalysis2012
provides strong empirical support: the summary measure approach
is robust to covariance misspecification, whereas LMM can
produce unreliable results when the working covariance
structure is wrong.

The practical advantages of summary statistics are not merely
aesthetic. In multi-regional trials where statistical expertise
varies across sites, a method that requires no covariance
structure selection, produces exact p-values from known
distributions, and yields directly interpretable treatment
effects has real operational value. The
degrees-of-freedom controversy surrounding
@kenwardSmallSampleInference1997 approximations is not a
theoretical curiosity: different software implementations
produce different p-values, creating regulatory risk when
results from SAS and R disagree.

@sennRepeatedMeasuresClinical2000 made the important
observation that summary measures preserve
intention-to-treat principles more naturally than some
repeated measures approaches. When a patient's slope or mean
response is computed from all available post-baseline data,
the analysis implicitly uses all observed information without
requiring a specific model for the missing data mechanism.
@dawsonStratificationSummaryStatistic1994 showed that
stratifying by missing data pattern further improves
robustness.

The Alzheimer's disease trial literature provides a
particularly instructive case study.
@ardMixedModelRepeated2011 and @ardSimulationStudyComparing2015
showed that slope models have a power advantage over MMRM when
disease progression is approximately linear, which is the
prevailing assumption for disease-modifying therapies. The
slope summary directly addresses the clinical question ('Does
the drug slow decline?') in a way that the MMRM treatment
effect at a single visit does not.

The limitations of summary statistics should be stated
plainly. When treatment effects are non-monotone (e.g., an
initial benefit that wanes), no single summary captures the
full trajectory, and MMRM's flexibility in modelling
visit-specific effects is genuinely advantageous. Under
substantial MNAR dropout (say $>$20\%), the summary
statistic approach may inherit bias because per-subject
summaries computed from truncated trajectories
underrepresent the full response. And when the scientific
question genuinely concerns the treatment effect at a
specific landmark time point (e.g., the FDA-mandated
co-primary endpoint at week 78 in Alzheimer's trials),
MMRM's ability to estimate visit-specific contrasts is
essential.

@liuMixedModelsCovariateInteractions2021 highlighted
another subtlety: MMRM achieves power gains over
complete-case ANCOVA only when time-by-covariate interactions
are included. Without these, MMRM may be no better than the
simpler method it purports to replace. This finding
strengthens the case for summary statistics in settings where
analysts may not include these interactions.

The estimands framework introduced by @ichE9R12019 provides
a useful lens. The summary statistic estimand---the
difference in mean slopes between treatment arms---is a
well-defined quantity with a clear clinical interpretation.
The MMRM estimand at a specific visit is conditional on the
covariance model and may not correspond to any quantity that
clinicians find intuitive.

# Future research

Several directions merit further investigation:

1. **Nonlinear summary statistics.** The present review
   focuses on linear summaries (slopes, means, change
   scores). Nonlinear summaries such as time to a clinically
   meaningful threshold or the area under the
   response-time curve (AUC) may capture treatment effects
   that linear summaries miss. Efficiency comparisons under
   realistic data-generating models would be valuable.

2. **Robust variance estimation.** Combining summary
   statistics with sandwich (HC) standard errors could
   provide robustness to heteroscedasticity without requiring
   mixed-model machinery. The properties of such estimators
   in the summary statistic context are unexplored.

3. **Multiple imputation with summary statistics.** When
   missing data rates are substantial, computing summary
   statistics within each multiply imputed dataset and pooling
   via Rubin's rules could extend the approach to settings
   where simple complete-case analysis is inadequate.

4. **Software tools.** A dedicated R package implementing the
   summary measures approach with ANCOVA, baseline adjustment,
   sample size formulas, and diagnostic plots would lower the
   barrier to adoption.

5. **Head-to-head regulatory submissions.** Empirical evidence
   from regulatory submissions where both MMRM and summary
   statistics were presented as co-primary or sensitivity
   analyses would illuminate the practical concordance of the
   two approaches.

6. **Bayesian summary statistics.** Integrating the summary
   measures approach with Bayesian sample size determination
   [@baghfalakiBayesianSampleSize2019] and posterior inference
   could provide a unified framework for design and analysis.

7. **Extension to non-Gaussian endpoints.** The summary
   measures framework is well-developed for continuous outcomes.
   Extension to binary or count outcomes (e.g., proportion of
   visits with response, total adverse events) remains
   underexplored.

# References
